{RILS}-{ROLS} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-12-03 paper_resources/doc/bmc_jbd/bmc_jbd.tex
Final polishing

In this work we present a novel approach to solve \emph{ground-truth}-based SR, which combines the popular iterated local search meta-heuristic (ILS)~\cite{lourencco2003iterated,lourencco2019iterated} with the ordinary least square method (OLS)~\cite{leng2007ordinary}. ILS mostly handles combinatorial (discrete) aspects of search space, while OLS helps in the process of a coefficient determination, so it handles some continuous parts of the search space. As will be shown later, the proposed method has shown its robustness w.r.t. introduced noise, therefore we called the method \textsc{RILS}-\textsc{ROLS}  (with letters R corresponding to regression and robust). Additionally, in order to navigate the search toward exact model (and not only accurate one) the algorithm is equipped with a carefully constructed fitness function that combines three important model characteristics: RMSE and $R^2$ scores which participate into models accuracy, and a weighted solution size that penalizes too complex models.

@@

\begin{table}[!h]
	\caption{Comparison without noise}\label{tab:comp_noise0}
	\centering
	\begin{tabular}{l|rrr|rrr} \hline
		& \multicolumn{3}{c|}{Exact model percentage} & \multicolumn{3}{c}{$R^2 > 0.999$ percentage}\\ \hline
		Method & Feynman & Strogatz & Total & Feynman & Strogatz & Total \\ \hline
		\textsc{RILS-ROLS}&\bf{57.84}&\bf{83.57}&\bf{60.62}&82.59&90&83.38\\
		\textsc{AI-Feynman}&55.78&27.14&52.65&78.51&35.71&73.83\\
		\textsc{GP-GOMEA}&26.83&29.46&27.12&71.55&71.43&71.54\\
		\textsc{AFP-FE}&26.98&20&26.23&59.05&28.57&55.77\\
		\textsc{ITEA}&22.41&7.14&20.77&27.59&21.43&26.92\\
		\textsc{AFP}&21.12&15.18&20.48&44.83&25&42.69\\
		\textsc{DSR}&19.72&19.64&19.71&25&14.29&23.85\\
		\textsc{Operon}&16.55&11.43&16&86.21&\bf{92.86}&86.92\\
		\textsc{gplearn}&16.27&8.93&15.48&32.76&7.14&30\\
		\textsc{SBP-GP}&12.72&11.61&12.6&73.71&78.57&74.23\\
		\textsc{EPLEX}&12.39&8.93&12.02&46.98&21.43&44.23\\
		\textsc{BSR}&2.48&0.89&2.31&10.78&21.43&11.92\\
		\textsc{FEAT}&0&0.89&0.1&39.66&42.86&40\\
		\textsc{FFX}&0&0&0&0&0&0\\
		\textsc{MRGP}&0&0&0&\bf{93.1}&89.29&\bf{92.69}\\
		\hline
	\end{tabular}
\end{table}

\begin{table}[!h]
	\caption{Comparison with noise level 0.001}\label{tab:comp_noise0001}
	\centering
	\begin{tabular}{l|rrr|rrr} \hline
		& \multicolumn{3}{c|}{Exact model percentage} & \multicolumn{3}{c}{$R^2 > 0.999$ percentage}\\ \hline
		Method & Feynman & Strogatz & Total & Feynman & Strogatz & Total \\ \hline
		\textsc{RILS-ROLS}&\bf{42.93}&\bf{35}&\bf{42.08}&80.26&65&78.62\\
		\textsc{AI-Feynman}&33.08&22.14&31.89&77.39&42.86&73.64\\
		\textsc{AFP-FE}&21.9&15.71&21.23&52.16&35.71&50.38\\
		\textsc{DSR}&19.14&19.29&19.15&25.86&14.29&24.62\\
		\textsc{AFP}&19.66&13.57&19&44.4&21.43&41.92\\
		\textsc{gplearn}&16.99&9.29&16.16&31.9&7.14&29.23\\
		\textsc{ITEA}&14.57&7.14&13.77&27.59&21.43&26.92\\
		\textsc{Operon}&13.19&5&12.31&85.78&\bf{92.86}&86.54\\
		\textsc{GP-GOMEA}&11.03&7.14&10.62&70.26&71.43&70.38\\
		\textsc{EPLEX}&9.57&9.29&9.54&47.84&25&45.38\\
		\textsc{SBP-GP}&0.78&0&0.69&75.43&64.29&74.23\\
		\textsc{BSR}&0.6&0.71&0.62&10.34&14.29&10.77\\
		\textsc{FFX}&0&0&0&0&0&0\\
		\textsc{FEAT}&0&0&0&42.24&50&43.08\\
		\textsc{MRGP}&0&0&0&\bf{92.24}&85.71&\bf{91.54}\\
		\hline
	\end{tabular}
\end{table}

\begin{table}[!h]
	\caption{Comparison with noise level 0.01}\label{tab:comp_noise001}
	\centering
	\begin{tabular}{l|rrr|rrr} \hline
		& \multicolumn{3}{c|}{Exact model percentage} & \multicolumn{3}{c}{$R^2 > 0.999$ percentage}\\ \hline
		Method & Feynman & Strogatz & Total & Feynman & Strogatz & Total \\ \hline
		\textsc{RILS-ROLS}&\bf{36.29}&\bf{22.14}&\bf{34.77}&64.91&46.43&62.92\\
		\textsc{AFP-FE}&20.78&13.57&20&52.59&32.14&50.38\\
		\textsc{DSR}&18.97&18.57&18.92&26.29&14.29&25\\
		\textsc{AFP}&16.9&11.43&16.31&42.67&21.43&40.38\\
		\textsc{gplearn}&16.87&9.29&16.05&29.13&10.71&27.13\\
		\textsc{AI-Feynman}&13.03&9.29&12.61&71.43&39.29&67.86\\
		\textsc{EPLEX}&8.71&4.29&8.23&56.03&21.43&52.31\\
		\textsc{ITEA}&7.84&6.43&7.69&27.59&21.43&26.92\\
		\textsc{GP-GOMEA}&5.09&1.43&4.69&73.71&71.43&73.46\\
		\textsc{Operon}&2.07&0.71&1.92&85.78&\bf{92.86}&86.54\\
		\textsc{BSR}&0.09&0&0.08&12.07&10.71&11.92\\
		\textsc{SBP-GP}&0&0&0&75&75&75\\
		\textsc{FEAT}&0&0&0&40.52&42.86&40.77\\
		\textsc{FFX}&0&0&0&2.59&3.57&2.69\\
		\textsc{MRGP}&0&0&0&\bf{91.81}&60.71&\bf{88.46}\\
		\hline
	\end{tabular}
\end{table}

@@

\emph{Important note. Exact model accuracy of \textsc{RILS-ROLS} is calculated by dividing the number of successful runs with 1300, i.e., the maximal possible number of successful runs. 
On the other hand, the exact model accuracy of methods revised in \texttt{SRBench} was calculated in a different way. First, the accuracy of each of 130 considered instances was calculated by dividing the number of successful runs with the number of runs that finished its execution. This means that, for example, if some instance was run 10 times, giving 4 successful models, but failing to finish 2 runs due to memory exhaustion or some other OS issue, the accuracy would be 50\% (4/8) instead of 40\% (4/10). Further, these individual accuracies were averaged. 
We do not believe this was the right way to calculate the accuracy, because, this allows one to have high-accuracy method with a low number of successful runs. For example, if method A finishes all of its runs, out of which 650 are successful, its total accuracy will be 50\%. On the other hand, one can have a method B that finishes instances execution in only 4 out of 10 runs, giving 2 successful models each time -- its per-instances accuracy will be 50\%, and after averaging across all instances it will be also 50\%. So, methods A and B will have the same overall accuracy, although method A reached 650 correct solutions, while method B reached only 260. Also, note that our calculation cannot overestimate \textsc{RILS-ROLS} exact model accuracy, since denominator is always maximal (1300) -- in essence, non-finished runs are counted as non-successful.}

@@

The following conclusions may be drawn from these results. \fxnote{Marko: dosao dovde!}

@@

\begin{itemize}
	\item For data without noise, the exact percentage of \textsc{RILS}-\textsc{ROLS} slightly decreases as the  number of variables increases. 
	\item The exact percentages drop off with the presence of a larger noise level. As expected, higher noise means smaller exact percentage.
	\item Interestingly, in the presence of noise in data, the highest exact percentages are not obtained for the instances with a single input variable, but for those with two or three. This is probably due to the expected level of \emph{non-linearity} in instances. Namely, when the size of instance is fixed, instances with less input variables will have to use more operations to compensate for the lack of input variables. (Note that expression size is calculated simply by counting all nodes in the expression tree, whether these nodes are operations, constants or variables.) Therefore, instances with less variables will more heavily use operations and their compositions, paving a way to more \emph{difficult} functional structure. 
	The reason this is not an issue for no-noisy data is because \textsc{RILS-ROLS} will have to wander less -- there will be no contradictory search directions as with noisy data.
\end{itemize}

@@

\section*{Acknowledgements}%% if any
Text for this section\ldots

\section*{Funding}%% if any
Text for this section\ldots

\section*{Abbreviations}%% if any
Text for this section\ldots

\section*{Availability of data and materials}%% if any
Text for this section\ldots

\section*{Ethics approval and consent to participate}%% if any
Text for this section\ldots

\section*{Competing interests}
The authors declare that they have no competing interests.

\section*{Consent for publication}%% if any
Text for this section\ldots

\section*{Authors' contributions}
Text for this section \ldots

\section*{Authors' information}%% if any
Text for this section\ldots

@@

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure...

>>>>

In this work we present a novel approach to solve \emph{ground-truth}-based SR, which combines the popular iterated local search meta-heuristic (ILS)~\cite{lourencco2003iterated,lourencco2010iterated} with the ordinary least square method (OLS)~\cite{leng2007ordinary}. ILS mostly handles combinatorial (discrete) aspects of search space, while OLS helps in the process of a coefficient determination, so it handles some continuous parts of the search space. As will be shown later, the proposed method has shown its robustness w.r.t. introduced noise, therefore we called the method \textsc{RILS}-\textsc{ROLS}  (with letters R corresponding to regression and robust). Additionally, in order to navigate the search toward exact model (and not only accurate one) the algorithm is equipped with a carefully constructed fitness function that combines three important model characteristics: RMSE and $R^2$ scores which participate into models accuracy, and a weighted solution size that penalizes too complex models.

@@

\begin{table}[!h]
	\caption{Comparison without noise}\label{tab:comp_noise0}
	\centering
	\begin{tabular}{l|rrr|rrr} \hline
		& \multicolumn{3}{c|}{Exact model percentage} & \multicolumn{3}{c}{$R^2 > 0.999$ percentage}\\ \hline
		Method &  \textsc{Feynman} & \textsc{Strogatz}& Total &  \textsc{Feynman} & \textsc{Strogatz} & Total \\ \hline
		\textsc{RILS-ROLS}&\bf{57.84}&\bf{83.57}&\bf{60.62}&82.59&90&83.38\\
		\textsc{AI-Feynman}&55.78&27.14&52.65&78.51&35.71&73.83\\
		\textsc{GP-GOMEA}&26.83&29.46&27.12&71.55&71.43&71.54\\
		\textsc{AFP-FE}&26.98&20&26.23&59.05&28.57&55.77\\
		\textsc{ITEA}&22.41&7.14&20.77&27.59&21.43&26.92\\
		\textsc{AFP}&21.12&15.18&20.48&44.83&25&42.69\\
		\textsc{DSR}&19.72&19.64&19.71&25&14.29&23.85\\
		\textsc{Operon}&16.55&11.43&16&86.21&\bf{92.86}&86.92\\
		\textsc{gplearn}&16.27&8.93&15.48&32.76&7.14&30\\
		\textsc{SBP-GP}&12.72&11.61&12.6&73.71&78.57&74.23\\
		\textsc{EPLEX}&12.39&8.93&12.02&46.98&21.43&44.23\\
		\textsc{BSR}&2.48&0.89&2.31&10.78&21.43&11.92\\
		\textsc{FEAT}&0&0.89&0.1&39.66&42.86&40\\
		\textsc{FFX}&0&0&0&0&0&0\\
		\textsc{MRGP}&0&0&0&\bf{93.1}&89.29&\bf{92.69}\\
		\hline
	\end{tabular}
\end{table}

\begin{table}[!h]
	\caption{Comparison with noise level 0.001}\label{tab:comp_noise0001}
	\centering
	\begin{tabular}{l|rrr|rrr} \hline
		& \multicolumn{3}{c|}{Exact model percentage} & \multicolumn{3}{c}{$R^2 > 0.999$ percentage}\\ \hline
		Method &  \textsc{Feynman} & \textsc{Strogatz} & Total &  \textsc{Feynman} & \textsc{Strogatz} & Total \\ \hline
		\textsc{RILS-ROLS}&\bf{42.93}&\bf{35}&\bf{42.08}&80.26&65&78.62\\
		\textsc{AI-Feynman}&33.08&22.14&31.89&77.39&42.86&73.64\\
		\textsc{AFP-FE}&21.9&15.71&21.23&52.16&35.71&50.38\\
		\textsc{DSR}&19.14&19.29&19.15&25.86&14.29&24.62\\
		\textsc{AFP}&19.66&13.57&19&44.4&21.43&41.92\\
		\textsc{gplearn}&16.99&9.29&16.16&31.9&7.14&29.23\\
		\textsc{ITEA}&14.57&7.14&13.77&27.59&21.43&26.92\\
		\textsc{Operon}&13.19&5&12.31&85.78&\bf{92.86}&86.54\\
		\textsc{GP-GOMEA}&11.03&7.14&10.62&70.26&71.43&70.38\\
		\textsc{EPLEX}&9.57&9.29&9.54&47.84&25&45.38\\
		\textsc{SBP-GP}&0.78&0&0.69&75.43&64.29&74.23\\
		\textsc{BSR}&0.6&0.71&0.62&10.34&14.29&10.77\\
		\textsc{FFX}&0&0&0&0&0&0\\
		\textsc{FEAT}&0&0&0&42.24&50&43.08\\
		\textsc{MRGP}&0&0&0&\bf{92.24}&85.71&\bf{91.54}\\
		\hline
	\end{tabular}
\end{table}

\begin{table}[!h]
	\caption{Comparison with noise level 0.01}\label{tab:comp_noise001}
	\centering
	\begin{tabular}{l|rrr|rrr} \hline
		& \multicolumn{3}{c|}{Exact model percentage} & \multicolumn{3}{c}{$R^2 > 0.999$ percentage}\\ \hline
		Method & \textsc{Feynman} & \textsc{Strogatz} & Total & \textsc{Feynman} & \textsc{Strogatz} & Total \\ \hline
		\textsc{RILS-ROLS}&\bf{36.29}&\bf{22.14}&\bf{34.77}&64.91&46.43&62.92\\
		\textsc{AFP-FE}&20.78&13.57&20&52.59&32.14&50.38\\
		\textsc{DSR}&18.97&18.57&18.92&26.29&14.29&25\\
		\textsc{AFP}&16.9&11.43&16.31&42.67&21.43&40.38\\
		\textsc{gplearn}&16.87&9.29&16.05&29.13&10.71&27.13\\
		\textsc{AI-Feynman}&13.03&9.29&12.61&71.43&39.29&67.86\\
		\textsc{EPLEX}&8.71&4.29&8.23&56.03&21.43&52.31\\
		\textsc{ITEA}&7.84&6.43&7.69&27.59&21.43&26.92\\
		\textsc{GP-GOMEA}&5.09&1.43&4.69&73.71&71.43&73.46\\
		\textsc{Operon}&2.07&0.71&1.92&85.78&\bf{92.86}&86.54\\
		\textsc{BSR}&0.09&0&0.08&12.07&10.71&11.92\\
		\textsc{SBP-GP}&0&0&0&75&75&75\\
		\textsc{FEAT}&0&0&0&40.52&42.86&40.77\\
		\textsc{FFX}&0&0&0&2.59&3.57&2.69\\
		\textsc{MRGP}&0&0&0&\bf{91.81}&60.71&\bf{88.46}\\
		\hline
	\end{tabular}
\end{table}

@@

\emph{Important note. Exact model accuracy of \textsc{RILS-ROLS} is calculated by dividing the number of successful runs with 1300, i.e., the maximal possible number of successful runs. 
On the other hand, the exact model accuracy of methods revised in \texttt{SRBench} was calculated in a different way. First, the accuracy of each of 130 considered instances was calculated by dividing the number of successful runs with the number of runs that finished its execution. This means that, for example, if some instance was run 10 times, giving 4 successful models, but failing to finish 2 runs due to memory exhaustion or some other OS issue, the accuracy would be 50\% (4/8) instead of 40\% (4/10). Further, these individual accuracies were averaged. 
We do not believe this was the right way to calculate the accuracy, because, this allows one to have high-accuracy method with a low number of successful runs. For example, if method A finishes all of its runs, out of which 650 are successful, its total accuracy will be 50\%. On the other hand, one can have a method B that finishes instances execution in only 4 out of 10 runs, giving 2 successful models each time -- its per-instances accuracy will be 50\%, and after averaging across all instances it will be also 50\%. So, methods A and B will have the same overall accuracy, although method A reached 650 correct solutions, while method B reached only 260.}

@@

The following conclusions may be drawn from these results.

@@

\begin{itemize}
	\item For data without noise, the exact percentage of \textsc{RILS}-\textsc{ROLS} slightly decreases as the  number of variables increases. 
	\item The exact percentages drop off with the presence of a larger noise level. As expected, higher noise means smaller exact percentage.
	\item Interestingly, in the presence of noise in data, the highest exact percentages are not obtained for the instances with a single input variable, but for those with two or three. This is probably due to the expected level of \emph{non-linearity} in instances. Namely, when the size of instance is fixed, instances with less input variables will have to use more operations to compensate for the lack of input variables. (Note that expression size is calculated simply by counting all nodes in the expression tree, whether these nodes are operations, constants or variables.) Therefore, instances with less variables will more heavily use operations and their compositions, paving a way to more \emph{difficult} functional structure. 
	The reason this is not an issue for no-noisy data is because \textsc{RILS-ROLS} will have to wander less -- there will be no so many contradictory search directions as with noisy data.
\end{itemize}

@@

 \section*{Acknowledgments}%% if any
Not applicable.

 \section*{Funding}%% if any
Not applicable.

%\section*{Abbreviations}%% if any
%Text for this section\ldots

  \section*{Availability of data and materials}%% if any
 All accompanying resources regarding this paper can be found under the repository \url{https://github.com/kartelj/rils-rols}.

\section*{Ethics approval and consent to participate}%% if any
Not applicable.

\section*{Competing interests}
The authors declare  no competing interests.

\section*{Consent for publication}%% if any
Not applicable.

\section*{Authors' contributions}
A.K. and M.D. contributed to the conception of the work. A.K. worked on implementing the algorithm and the design of experiments. M.D. worked on analysis and discussion of the results. Both authors read and approved the final manuscript.

%\section*{Authors' information}%% if any
%Text for this section\ldots

@@

%\section*{Additional Files}
%  \subsection*{Additional file 1 --- Sample additional file title}
 %   Additional file descriptions text (including details of how to
 %   view the file, if it is in a non-standard format or the file extension).  This might
 %   refer to a multi-page table or a figure...
